---
title: Blog Post 7 -- The Ground Game
author: Chris S
date: '2024-10-21'
slug: blog-post-7-the-ground-game
categories: []
tags: []
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(geofacet)
library(ggpubr)
library(ggthemes)
library(haven)
library(kableExtra)
library(maps)
library(mgcv)
library(mgcViz)
library(RColorBrewer)
library(scales)
library(sf)
library(spData)
library(stargazer)
library(tidygeocoder)
library(tidyverse)
library(tigris)
library(tmap)
library(tmaptools)
library(viridis)
```


```{r, echo = FALSE, message = FALSE, warning = FALSE}
d_popvote <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/popvote_1948-2020.csv")
d_state_popvote <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/state_popvote_1948_2020.csv")
d_state_popvote[d_state_popvote$state == "District of Columbia",]$state <- "District Of Columbia"

# Read elector distribution dataset. 
d_ec <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/corrected_ec_1948_2024.csv")

# Read polling data. 
d_polls <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/national_polls_1968-2024.csv")
d_state_polls <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/state_polls_1968-2024.csv")

# Read turnout data. 
d_turnout <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/state_turnout_1980_2022.csv")

# Read county turnout. 
d_county_turnout <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/county_turnout.csv")

# Read state-level demographics.
d_state_demog <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/demographics.csv")

# Read county demographics. 
d_county_demog <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/county_demographics.csv")

# Read campaign events datasets. 
d_campaign_events <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/campaigns_2016_2024.csv")[,-1]
```





*This blog is an ongoing assignment for Gov 1347: Election Analytics, a course at Harvard College taught by Professor [Ryan Enos](https://www.ryandenos.com/). It will be updated weekly and culminate in a predictive model of the 2024 presidential election.*

## Introduction -- What role does the *VEP* play in state-level election results? How effective is the *ground game* at swinging election outcomes? Ultimately, can *campaign events* help predict these results?

*Knock knock.* You open your front door to find a **campaign volunteer**, clipboard in hand, eager to talk to you about the upcoming election. We've all seen these people randomly roaming our neighborhoods. In reality, this is **just one piece of the larger ground game strategy,** where candidates and their teams hit the pavement, host rallies, and knock on doors to **sway voters face-to-face.** But with digital ads and viral tweets dominating the outreach scene, **how effective is this old-school approach?**

Some argue that this holds the secret to electoral success, while others say that rallies and door-knocking are **more symbolic than substantial.** So, can these campaign events and personal interactions **really move the needle** and predict state-level results? Or, is this all **just political theater?**

As always, let's start with the data!

## Back To Polls?

To begin our analysis, let's take a step back and first examine **state-specific forecasts for Democratic and Republican vote shares** based on polling data:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Merge popular vote and polling data. 
d <- d_state_popvote |> 
  inner_join(d_state_polls |> filter(weeks_left == 3)) |> 
  mutate(state_abb = state.abb[match(state, state.name)])

# Generate state-specific univariate poll-based forecasts with linear model.
state_forecast <- list()
state_forecast_outputs <- data.frame()
for (s in unique(d$state_abb)) {
  # Democrat model.
  state_forecast[[s]]$dat_D <- d |> filter(state_abb == s, party == "DEM")
  state_forecast[[s]]$mod_D <- lm(D_pv ~ poll_support, 
                                  state_forecast[[s]]$dat_D)
  
  # Republican model.
  state_forecast[[s]]$dat_R <- d |> filter(state_abb == s, party == "REP")
  state_forecast[[s]]$mod_R <- lm(R_pv ~ poll_support, 
                                  state_forecast[[s]]$dat_R)
  
  if (nrow(state_forecast[[s]]$dat_R) > 2) {
    # Save state-level model estimates. 
    state_forecast_outputs <- rbind(state_forecast_outputs, 
                                    rbind(cbind.data.frame(
                                      intercept = summary(state_forecast[[s]]$mod_D)$coefficients[1,1], 
                                      intercept_se = summary(state_forecast[[s]]$mod_D)$coefficients[1,2],
                                      slope = summary(state_forecast[[s]]$mod_D)$coefficients[2,1], 
                                      state_abb = s, 
                                      party = "DEM"), 
                                    rbind(cbind.data.frame(
                                     intercept = summary(state_forecast[[s]]$mod_R)$coefficients[1,1],
                                     intercept_se = summary(state_forecast[[s]]$mod_R)$coefficients[1,2],
                                     slope = summary(state_forecast[[s]]$mod_R)$coefficients[2,1],
                                     state_abb = s,
                                     party = "REP"
                                    ))))
  }
}

# Make graphs of polls in different states/parties at different levels of strength/significance of outcome. 
state_forecast_trends <- state_forecast_outputs |> 
  mutate(`0` = intercept, 
         `25` = intercept + slope*25, 
         `50` = intercept + slope*50, 
         `75` = intercept + slope*75, 
         `100` = intercept + slope*100) |>
  select(-intercept, -slope) |> 
  gather(x, y, -party, -state_abb, -intercept_se) |> 
  mutate(x = as.numeric(x))

# Q: What's wrong with this map? 
# A: (1.) no polls in some states
#    (2.) very high variance for some states (Nevada)/negative slopes for others (Mississippi)
#    (3.) y is not always in the [0, 100] range
ggplot(state_forecast_trends, aes(x=x, y=y, ymin=y-intercept_se, ymax=y+intercept_se)) + 
  facet_geo(~ state_abb) +
  geom_line(aes(color = party)) + 
  geom_ribbon(aes(fill = party), alpha=0.5, color=NA) +
  coord_cartesian(ylim=c(0, 100)) +
  scale_color_manual(values = c("blue", "red")) +
  scale_fill_manual(values = c("blue", "red")) +
  xlab("Hypothetical Poll Support") +
  ylab("Predicted Voteshare\n(pv = A + B * poll)") +
  ggtitle("") +
  theme_bw()
```

Here, we can see that for each state, **the blue line represents the Democratic vote share prediction**, while the **red line** represents the **Republican vote share.** Immediately, we notice that for most states, **as poll support increases,** the predicted vote share for **both parties also increases.** This makes sense. 

However, there are also some issues with the plot. For example, some states have odd patterns, **such as negative slopes or very high variance,** suggesting that the linear models **may not fit well** in certain states. For example, in **Nevada (NV), we see high variance in predictions, while in Mississippi (MS),** there might be a negative slope, which is **unrealistic** for real-world elections. Historically, polling errors are more pronounced in states with **smaller or harder-to-reach populations,** such as Nevada. Other accuracy concerns include factors like **partisan lean and polling methodology.**

Thus, we're seeing instances when polling data might be **unreliable** or where the relationship between polling support and actual vote share is **rather weak.**

Let's take a closer look at two key states:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
state_forecast_trends |>
  filter(state_abb == "CA" | state_abb == "FL")|>
  ggplot(aes(x=x, y=y, ymin=y-intercept_se, ymax=y+intercept_se)) + 
  facet_wrap(~ state_abb) +
  geom_line(aes(color = party)) + 
  geom_hline(yintercept = 100, lty = 3) +
  geom_hline(yintercept = 0, lty = 3) + 
  geom_ribbon(aes(fill = party), alpha=0.5, color=NA) +
  ## N.B. You can, in fact, combine *different* data and aesthetics
  ##       in one ggplot; but this usually needs to come at the end 
  ##       and you must explicitly override all previous aesthetics
  geom_text(data = d |> filter(state_abb == "CA", party=="DEM"), 
            aes(x = poll_support, y = D_pv, ymin = D_pv, ymax = D_pv, color = party, label = year), size=1.5) +
  geom_text(data = d |> filter(state_abb == "CA", party=="REP"), 
            aes(x = poll_support, y = D_pv, ymin = D_pv, ymax = D_pv, color = party, label = year), size=1.5) +
  geom_text(data = d |> filter(state_abb == "FL", party=="DEM"), 
            aes(x = poll_support, y = D_pv, ymin = D_pv, ymax = D_pv, color = party, label = year), size=1.5) +
  geom_text(data = d |> filter(state_abb == "FL", party=="REP"), 
            aes(x = poll_support, y = D_pv, ymin = D_pv, ymax = D_pv, color = party, label = year), size=1.5) +
  scale_color_manual(values = c("blue", "red")) +
  scale_fill_manual(values = c("blue", "red")) +
  xlab("Hypothetical Poll Support") +
  ylab("Predicted Voteshare\n(pv = A + B * poll)") +
  theme_bw()
```

For California, we can see that the **Democrats have a higher predicted vote share across almost all polling support levels** compared to Republicans. Historically democratic years like **2020, 2016, and 2008 align closely** with the predicted vote share for Democrats at mid-to-high polling support. On the other hand, years like 1980 and 1984 show Republicans doing better, but this **makes sense as California was less reliably blue at the time.**

In contrast, Florida reveals how the lines are closer together, indicating a **more competitive electoral environment** between the two parties. At the same time, the confidence intervals in Florida are wider, reflecting **more uncertainty** in the model's predictions. This is likely due to Florida’s **more volatile voting patterns** and demographic shifts, making it harder to predict election outcomes based solely on polling.

## Voting Eligible Population -- Pennsylvania

Let's take a deeper dive into the **voting-eligible population (VEP).** These people represent the portion of the population that is **legally eligible to vote.** Knowing the VEP helps in predicting election outcomes, **especially in close races.** A growing or shrinking VEP can significantly **alter electoral dynamics, especially in battleground states.**

Let's examine a specific battleground case study:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Simulating a distribution of potential election results in Pennsylvania for 2024. 
# First step. Let's use GAM (general additive model) to impute VEP in Pennsylvania for 2024 using historical VEP.

# Get historical eligible voting population in Pennsylvania. 
vep_PA_2020 <- as.integer(d_turnout$vep[d_turnout$state == "Pennsylvania" & d_turnout$year == 2020])
vep_PA <- d_turnout |> filter(state == "Pennsylvania") |> select(vep, year)

# Fit regression for 2024 VEP prediction. 
lm_vep_PA <- lm(vep ~ year, vep_PA)

plot(x = vep_PA$year, y = vep_PA$vep, xlab = "Year", ylab = "VEP", main = "Voting Eligible Population in Pennsylvania by Year")
abline(lm_vep_PA, col = "red")

```

Here, we see that Pennsylvania has shown a **steady, nearly linear increase in the VEP** over the last few decades, reflecting population growth and demographic shifts. The red regression line hugs the points closely, **predicting future VEP based on historical trends.** Thus, we can ultimately assess how much influence a growing population could have on **turnout rates and on election results.**

Now, let's expand the OLS model to include **generalized additive modeling (GAM)** and use a weighted average to create a final prediction:

```{r, echo = FALSE, message = FALSE, warning = FALSE}

vep_PA_2024_ols <- predict(lm_vep_PA, newdata = data.frame(year = 2024)) |> as.numeric()

gam_vep_PA <- mgcv::gam(vep ~ s(year), data = vep_PA)
print(plot(getViz(gam_vep_PA)) + l_points() + l_fitLine(linetype = 3) + l_ciLine(colour = 2) + theme_get()) 

# Use generalized additive model (GAM) to predict 2024 VEP in Pennsylvania.
vep_PA_2024_gam <- predict(gam_vep_PA, newdata = data.frame(year = 2024)) |> as.numeric()

# Take weighted average of linear and GAM predictions for final prediction. 
vep_PA_2024 <- as.integer(0.75*vep_PA_2024_gam + 0.25*vep_PA_2024_ols)
vep_PA_2024
```

Here, we're adding a **75% weight to the GAM prediction, and a 25% weight to the OLS prediction.** We can immediately see how GAM provides much **more flexibility** by allowing the model to fit a **curved relationship between VEP and time,** capturing complex demographic changes that might influence VEP growth in the future. As a result, **our final VEP prediction for 2024 is approximately, 10,044,706.**

Crossing 10 million eligible voters by the next election has **significant implications** for Pennsylvania's electoral dynamics, as a larger electorate could **change the composition of young voters,** influencing overall turnout and election outcomes.

## Field Offices and The Ground Game

Field offices play a crucial role in elections in terms of **voter outreach and mobilization.** These offices act as hubs for campaigns to **connect directly with voters, organize events, and more.** This allows for more specialized campaign strategies with tailored messaging and **more rapid response to local issues.**

Let's trace the data across 2012 and 2016:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Where should campaigns build field offices? 
options(readr.show_progress = FALSE)
fo_2012 <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/fieldoffice_2012_bycounty.csv")

lm_obama <- lm(obama12fo ~ romney12fo + 
                 swing + 
                 core_rep + 
                 swing:romney12fo + 
                 core_rep:romney12fo + 
                 battle + 
                 medage08 + 
                 pop2008 + 
                 pop2008^2 + 
                 medinc08 + 
                 black + 
                 hispanic + 
                 pc_less_hs00 + 
                 pc_degree00 + 
                 as.factor(state), 
               fo_2012)

lm_romney <- lm(romney12fo ~ 
                  obama12fo + 
                  swing + 
                  core_dem + 
                  swing:obama12fo + 
                  core_dem:obama12fo + 
                  battle + 
                  medage08 + 
                  pop2008 + 
                  pop2008^2 + 
                  medinc08 + 
                  black + 
                  hispanic + 
                  pc_less_hs00 + 
                  pc_degree00 + 
                  as.factor(state),
                  fo_2012)

# Effects of field offices on turnout and vote share. 
fo_dem <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/fieldoffice_2004-2012_dems.csv")

ef_t <- lm(turnout_change ~ dummy_fo_change + battle + dummy_fo_change:battle + as.factor(state) + as.factor(year), fo_dem)

ef_d <- lm(dempct_change ~ dummy_fo_change + battle + dummy_fo_change:battle + as.factor(state) + as.factor(year), fo_dem)

# Field Strategies of Obama, Romney, Clinton, and Trump in 2016. 
fo_add <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/fieldoffice_2012-2016_byaddress.csv")

invisible(us_geo <- counties(cb = TRUE) |> 
  shift_geometry() |> 
  filter(STUSPS %in% unique(fo_add$state)))

obama12 <- subset(fo_add, year == 2012 & candidate == "Obama") %>%
  select(longitude, latitude)
romney12 <- subset(fo_add, year == 2012 & candidate == "Romney") %>%
  select(longitude, latitude)
clinton16 <- subset(fo_add, year == 2016 & candidate == "Clinton") %>%
  select(longitude, latitude)
trump16 <- subset(fo_add, year == 2016 & candidate == "Trump") %>%
  select(longitude, latitude)

obama12transformed <- st_as_sf(obama12, coords = c("longitude", "latitude"), crs = 4326) |> 
  st_transform(crs = st_crs(us_geo)) |> 
  shift_geometry()
romney12transformed <- st_as_sf(romney12, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(us_geo)) |>
  shift_geometry()
clinton16transformed <- st_as_sf(clinton16, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(us_geo)) |>
  shift_geometry()
trump16transformed <- st_as_sf(trump16, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(us_geo)) |>
  shift_geometry()

invisible((ob12 <- ggplot() +
  geom_sf(data = us_geo) + 
  geom_sf(data = obama12transformed, color = "dodgerblue4", alpha = 0.75, pch = 3) +
  ggtitle("Obama 2012 Field Offices") +
  theme_void()))

invisible((ro12 <- ggplot() +
  geom_sf(data = us_geo) + 
  geom_sf(data = romney12transformed, color = "firebrick", alpha = 0.75, pch = 3) +
  ggtitle("Romney 2012 Field Offices") +
  theme_void()))

invisible((cl16 <- ggplot() + 
  geom_sf(data = us_geo) + 
  geom_sf(data = clinton16transformed, color = "dodgerblue4", alpha = 0.75, pch = 3) +
  ggtitle("Clinton 2016 Field Offices") +
  theme_void()))

invisible((tr16 <- ggplot() +
  geom_sf(data = us_geo) + 
  geom_sf(data = trump16transformed, color = "firebrick", alpha = 0.75, pch = 3) +
  ggtitle("Trump 2016 Field Offices") +
  theme_void()))

ggarrange(ob12, ro12, cl16, tr16, nrow = 2, ncol = 2)
```

Here, we can see how the Obama campaign **deployed a widespread and dense network of field offices,** with a heavy concentration in battleground states like **Ohio, Florida, and Pennsylvania.** Clinton followed a similar strategy, with a focus on the **Midwest battleground states**. However, a lagging voter turnout in crucial areas especially among **working-class voters in the Rust Belt** was one of the reasons for her defeat.

The Republicans showed a **significant shift from 2012 to 2016.** We can see how Trump's campaign had a more focused field office strategy, **targeting specific battlegrounds like Florida, Pennsylvania, Ohio, and Michigan.** However, his campaign relied heavily on **digital outreach and rallies** rather than a traditional field game. This unconventional method allowed him to increase his presence and reach, **generating a high turnout, particularly in rural and exurban areas.**

## The Democratic Ground Game

Let's dive deeper into Democratic field office work in battleground states by comparing some more historical trends:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Clinton 2016 Field Offices - Obama 2008 Field Offices. 
fo_add %>%
  subset(candidate %in% c("Clinton", "Obama") &
           state %in% c("CO", "FL", "IA", "MI", "NV", "NH", "NC", "OH", "PA", "VA", "WI")) %>%
  group_by(state, candidate) %>%
  summarize(fo = n()) %>%
  spread(key = candidate, value = fo) %>%
  mutate(diff = Clinton - Obama) %>%
  select(state, diff) %>%
  ggplot(aes(y = diff, x = state, fill = (diff > 0))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  ylim(-50, 15) +
  scale_y_continuous(breaks=seq(-50,10,10)) +
  xlab("State") +
  ylab("Clinton Field Offices - Obama '08 field offices")+
  theme_minimal() +
  theme(legend.position = "none",
        text = element_text(size = 15))
```

Here we can see how in most battleground states, **Clinton had fewer field offices in 2016 than Obama did in 2008,** indicated by the substantial negative values. This was detrimental when considering crucial states like Ohio, Florida, and Pennsylvania —- all critical swing states that contributed to Trump's victory.

Ultimately, we see that the graph shows how **Clinton's campaign may have underinvested in field offices** compared to Obama’s 2008 campaign, particularly in key Rust Belt states. These offices play a **crucial role in get-out-the-vote (GOTV) efforts,** volunteer mobilization, and localized campaign activities. This made it **more challenging to mobilize Democratic voters,** particularly in areas where **voter enthusiasm may have been lower** than in 2008.

Let's continue our analysis by looking at 2012 and 2016:

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# Case Study: Wisconsin Field Offices in 2012 and 2016 
obama12_wi <- subset(fo_add, year == 2012 & candidate == "Obama" & state == "WI") %>%
  select(longitude, latitude)
clinton16_wi <- subset(fo_add, year == 2016 & candidate == "Clinton" & state == "WI") %>%
  select(longitude, latitude)

obama12_wi_transformed <- st_as_sf(obama12_wi, coords = c("longitude", "latitude"), crs = 4326) |> 
  st_transform(crs = st_crs(us_geo)) |> 
  shift_geometry()
clinton16_wi_transformed <- st_as_sf(clinton16_wi, coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(us_geo)) |>
  shift_geometry()

obama_wi <- ggplot() +
  geom_sf(data = us_geo |> filter(STUSPS == "WI")) + 
  geom_sf(data = obama12_wi_transformed, color = "dodgerblue4", alpha = 0.75, pch=3, size=3, stroke=1) +
  ggtitle("Obama 2012 Field Offices in Wisconsin") +
  theme_void()

clinton_wi <- ggplot() +
  geom_sf(data = us_geo |> filter(STUSPS == "WI")) + 
  geom_sf(data = clinton16_wi_transformed, color = "dodgerblue4", alpha = 0.75, pch=3, size=3, stroke=1) +
  ggtitle("Clinton 2016 Field Offices in Wisconsin") +
  theme_void()

ggarrange(obama_wi, clinton_wi)
```

Once again, we see how Obama had a **dense network of field offices spread across Wisconsin,** particularly concentrated in urban and suburban areas, including the Milwaukee and Madison regions. His aggressive ground game aimed at **reaching a wide range of voters through traditional canvassing and mobilization efforts.** Comparatively, we can see Clinton's **fewer offices,** especially in rural areas and northern parts of the state.

```{r,echo = FALSE, message = FALSE, warning = FALSE}
pvcounty_wi <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/popvote_bycounty_2012-2016_WI.csv")
pv12_wi <- subset(pvcounty_wi, year == 2012) 
pv16_wi <- subset(pvcounty_wi, year == 2016) 

us_geo_12 <- us_geo |> 
  left_join(pv12_wi, by = c("NAME" = "county"))

us_geo_16 <- us_geo |>
  left_join(pv16_wi, by = c("NAME" = "county"))

obama_wis <- 
  ggplot() + 
  geom_sf(data = us_geo_12 |> filter(STUSPS == "WI"), aes(fill = D_win_margin)) + 
  geom_sf(data = obama12_wi_transformed, color = "dodgerblue4", alpha = 0.75, pch=3, size=3, stroke=1) +
  ggtitle("Obama 2012 Field Offices\nand Win Margin in Wisconsin") +
  theme_void() + 
  scale_fill_gradient2(high = "dodgerblue4", mid = "white", low = "firebrick1",
                       name = "Dem\nMargin") + 
  theme(plot.title = element_text(size=16, face="bold")) + theme(legend.position = "right")

clinton_wis <- 
  ggplot() + 
  geom_sf(data = us_geo_16 |> filter(STUSPS == "WI"), aes(fill = D_win_margin)) + 
  geom_sf(data = clinton16_wi_transformed, color = "dodgerblue4", alpha = 0.75, pch=3, size=3, stroke=1) +
  ggtitle("Clinton 2016 Field Offices\nand Win Margin in Wisconsin") +
  theme_void() + 
  scale_fill_gradient2(high = "dodgerblue4", mid = "white", low = "firebrick1",
                       name = "Dem\nMargin") + 
  theme(plot.title = element_text(size=16, face="bold")) + theme(legend.position = "right")


ggarrange(obama_wis, clinton_wis)
```

As expected, Obama **won many counties by comfortable margins,** particularly in urban areas like Milwaukee, Madison, and along the state's southern border. His distribution across both urban and rural areas helped **increase turnout in a wide range of regions.**

Comparatively for Clinton, many of the counties turned red, highlighting how **Trump outperformed Clinton in large parts of Wisconsin,** especially in rural areas. Her concentrated efforts in urban areas came at **the cost of losing support from rural and working-class voters.**

## So...What About 2024?

With the context and analysis we have now, let's examine current trends and events:

```{r, eval = FALSE, echo = FALSE, message = FALSE, warning = FALSE}

## DO NOT RUN AGAIN
d_campaign_events$party[d_campaign_events$candidate %in% c("Trump / Pence", "Trump", "Pence", "Trump/Pence", "Vance")] <- "REP"
d_campaign_events$party[d_campaign_events$candidate %in% c("Biden / Harris", "Biden", "Harris", "Biden/Harris", "Walz", "Kaine", "Clinton", "Clinton / Kaine")] <- "DEM"
p.ev.1 <- d_campaign_events |> group_by(date, party) |> summarize(n_events = n(), year) |> filter(year == 2016) |> ggplot(aes(x = date, y = n_events, color = party)) + geom_point() + geom_smooth() + ggtitle("2016") + theme_bw()
p.ev.2 <- d_campaign_events |> group_by(date, party) |> summarize(n_events = n(), year) |> filter(year == 2020) |> ggplot(aes(x = date, y = n_events, color = party)) + geom_point() + geom_smooth() + ggtitle("2020") +  theme_bw()
p.ev.3 <- d_campaign_events |> group_by(date, party) |> summarize(n_events = n(), year) |> filter(year == 2024) |> ggplot(aes(x = date, y = n_events, color = party)) + geom_point() + geom_smooth() + ggtitle("2024") + theme_bw()

ggarrange(p.ev.1, p.ev.2, p.ev.3)

# Mapping campaign events. 
d_campaign_events <- d_campaign_events |> 
  geocode(city = city, state = state, method = 'osm', lat = latitude , long = longitude)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
options(readr.show_progress = FALSE)
d_campaign_events <- read_csv("/Users/chrisshen/Downloads/GOV1347/Election Blog 4/campaign_events_geocoded.csv")
d_campaign_events$party[d_campaign_events$candidate %in% c("Trump / Pence", "Trump", "Pence", "Trump/Pence", "Vance")] <- "REP"
d_campaign_events$party[d_campaign_events$candidate %in% c("Biden / Harris", "Biden", "Harris", "Biden/Harris", "Walz", "Kaine", "Clinton", "Clinton / Kaine")] <- "DEM"

invisible(us_geo <- states(cb = TRUE) |> 
  shift_geometry() |> 
  filter(STUSPS %in% unique(fo_add$state)))

invisible(d_ev_transformed <- st_as_sf(d_campaign_events |> drop_na(), coords = c("longitude", "latitude"), crs = 4326) |>
  st_transform(crs = st_crs(us_geo)) |>
  shift_geometry() |> 
  st_make_valid())

invisible((ev16 <- ggplot() +
   geom_sf(data = us_geo) + 
   geom_sf(data = d_ev_transformed |> filter(year == 2016), aes(color = party), size = 3, alpha = 0.75, pch = 3) +
   ggtitle("2016 Campaign Events") +
   theme_void()))

invisible((ev20 <- ggplot() +
   geom_sf(data = us_geo) + 
   geom_sf(data = d_ev_transformed |> filter(year == 2020), aes(color = party), size = 3, alpha = 0.75, pch = 3) +
   ggtitle("2020 Campaign Events") +
   theme_void()))

(ev24 <- ggplot() +
   geom_sf(data = us_geo) + 
   geom_sf(data = d_ev_transformed |> filter(year == 2024), aes(color = party), size = 3, alpha = 0.75, pch = 3) +
   ggtitle("2024 Campaign Events") +
   theme_void())
```

As predicted, **both parties are concentrating their campaign events in battleground states like Michigan, Wisconsin, Pennsylvania, and Arizona.** These are crucial Rust Belt states that helped decide the 2020 election results. They also have a significant presence in Georgia and Arizona, **two historically Republican states that flipped to the Democrats in 2020,** suggesting that Democrats are working to consolidate their gains in these areas.

The Republicans are also concentrating in battleground states, but are also **hosting events in Ohio, Florida, and North Carolina, which have been traditionally red** but occasionally competitive. Doing so will help **secure support in historical strongholds** while trying to gain a **competitive edge in the swing states.**

Overall, we can see how both parties are hosting fewer campaign events in traditionally solid states, **allocating more of their resources to swing states where the outcomes are more unpredictable.**

## Campaign Events and Predicting Vote Share?

We can clearly see trends in campaign events, presence in battleground states, and historical results. **Thus, can we use this information to improve our model and predict vote share?**

Let's find out:

```{r,echo = FALSE, message = FALSE, warning = FALSE}
# Can the number of campaign events predict state-level vote share? 
d_ev_state <- d_campaign_events |> 
  group_by(year, state, party) |> 
  summarize(n_events = n()) |> 
  pivot_wider(names_from = party, values_from = n_events) |> 
  rename(n_ev_D = DEM, n_ev_R = REP)
d_ev_state$n_ev_D[which(is.na(d_ev_state$n_ev_D))] <- 0
d_ev_state$n_ev_R[which(is.na(d_ev_state$n_ev_R))] <- 0
d_ev_state$ev_diff_D_R <- d_ev_state$n_ev_D - d_ev_state$n_ev_R
d_ev_state$ev_diff_R_D <- d_ev_state$n_ev_R - d_ev_state$n_ev_D

d <- d |> 
  left_join(d_ev_state, by = c("year", "state_abb" = "state"))

lm_ev_D <- lm(D_pv2p ~ n_ev_D + ev_diff_D_R, data = d)

summary(lm_ev_D)
```

Here, we can see that for each additional Democratic event in a state, **the Democratic vote share is expected to increase by roughly 0.126 percentage points,** with a statistically significant p-value. However, the **R-squared value is very low,** suggesting that it only captures roughly 1.87% of the variance in state-level vote share based on the number of campaign events and the event difference.

Is this consistent with the Republican model?

```{r, echo = FALSE, message = FALSE, warning = FALSE}
lm_ev_R <- lm(R_pv2p ~ n_ev_R + ev_diff_R_D, data = d)
summary(lm_ev_R)
```

Interestingly, the model predicts that for each additional Republican event, **the Republican vote share decreases by roughly 0.126 percentage points**, with a statistically significant p-value. This could mean that Republicans tend to hold events in more competitive or difficult states, **where their baseline support is lower.**

However, the positive association in **ev_diff_R_D** suggests that Republicans **benefit more** then they hold more events than Democrats, showing how the ground game is **particularly effective when they out-campaign Democrats in terms of event presence.**

Still, the very low R-squared value suggests that while campaign events can influence voter engagement and media coverage, **they cannot account for most of the variance in state-level vote shares.** Statistically, the high residual standard error compared to the effect sizes indicates that there is **a lot of unexplained noise in the relationship between the predictors.** While the coefficients are statistically significant, meaning there is some measurable impact on vote share, **it is relatively small in the broader context of all factors** that determine election outcomes that we've seen in previous blogs.

Thus, I predict that its addition to our model will likely produce negligible impact...**especially when election day is only 14 days away!**

Stay tuned!

## Data Sources:

Data are from the GOV 1347 course materials. All files can be found [here](https://github.com/cys9772/election-blog4). All external sources are hyperlinked.
